---
title: "Bradley-Terry models in brms"
author: "Matthew Hoctor"
date: "2025-02-02"
date-modified: last-modified
license: "CC BY-SA 4.0"
categories: 
  - brms
  - Bayesian statistics
draft: false
freeze: true

editor: source
---

```{r setup}
#| include: false

# Load libraries

##    General:
library(tidyverse)
# library(data.table)
# library(devtools)
# library(remotes)
# library(lattice)                # alternate plotting system
# library(extraDistr)             # extra distributions
# library(philentropy)            # similarity & distance measures

##    Rethinking course:
# library(rethinking)             # generally don’t want to load rethinking and brms loaded simultaneously

## Directed Graphs
library(ggdag)
# library(dagitty)
library(igraph)
# library(rgl)                    # 3d graph visualization

##    Other Bayesian analysis packages:
# library(cmdstanr)               # loaded from rethinking package
# library(loo)                    # LOO-CV
# library(projpred)
# library(rstantools)

##    Kurtz packages
# library(ape)
# library(bayesplot)
library(brms)                   # BRMS
# library(broom)
# library(flextable)
library(GGally)                 # ggpairs
# library(ggdark)
# library(ggmcmc)
# library(ggrepel)
# library(ggthemes)
# library(ggtree)
# library(ghibli)
# library(gtools)                 #ddirichlet()
# library(invgamma)
# library(patchwork)
# library(posterior)
# library(psych)
# library(rcartocolor)
# library(Rcpp)
# library(rstan)
# library(santoku)
# library(StanHeaders)
# library(statebins)
library(tidybayes)              # mean_hdi()                
# library(viridis)
# library(viridisLite)
# library(wesanderson)

# Set options for cores, chains & threading:

cores = parallel::detectCores()
chains = ifelse(cores >=4, 4L, cores)
iter = 40000L
warmup = 2000L
options(
  mc.cores = parallel::detectCores(),
  # Within-chain parallelization; see vignette("brms_threading")
  brms.threads = threading(
    threads = cores %/% chains,       # == floor(cores/chains)
    grainsize = NULL,                 # default
    static = TRUE)                    # required for reproducibility; disables adaptive scheduling
)
# cleanup
rm(cores)
```

```{r UFClefties}
#| include: false

# load data from Statistical Rethinking course package:
data("UFClefties", package = "rethinking")
# write .csv file:
write.csv(UFClefties, file = "UFClefties.csv")
```

```{r fighters}
#| include: false

fighters <- rbind(
  UFClefties |> 
    select(fight, fighter1, fighter1.lefty, fighter1.win) |> 
    rename(fighter = fighter1,
           lefty = fighter1.lefty,
           win = fighter1.win),
  UFClefties |> 
    select(fight, fighter2, fighter2.lefty, fighter1.win) |> 
    rename(fighter = fighter2,
           lefty = fighter2.lefty,
           win = fighter1.win) |>
    mutate(win = if_else(win == 1, 0, 1))) |>
  group_by(fighter) |>
  mutate(
    Fights = length(fight),
    Wins = sum(win)
  ) |>
  select(-fight, -win) |>
  unique() |>
  ungroup() |>
  arrange(fighter)
```

## The UFClefties question

I recently came across an interesting question while working through the problem sets in the [Statistical Rethinking course](https://github.com/rmcelreath/stat_rethinking_2025).  This problem asks us to estimate the advantage of left-handedness within the UFC based on hand coded UFC data from [this paper](https://www.sciencedirect.com/science/article/abs/pii/S0003347213003485).  The question is motivated by the fact that lefties are over-represented in the UFC (~20% of fighters), as well as other sports like tennis, fencing, etc.  The [dataset](https://github.com/matthew-hoctor/blog/blob/main/posts/UFC/UFClefties.csv) includes info on each UFC fight in the years 2010-2012, including the winner and loser, and the handedness of each fighter

The solution set (and the paper) treat the abilities of each fighter as purely unobserved; i.e. the following causal structure is assumed for the data:

```{r DAG}
dagify(
  W ~ A + L + Q,
  Q ~ A + L,
  exposure = "L",
  outcome = "W",
  latent = "A",
  coords = list(
    x = c(L = 0, W = 1, A = 1,  Q = 0),
    y = c(L = 0, W = 0, A = 1,  Q = 1))) |>
  ggdag_status() + 
  theme_dag_gray_grid(
    panel.grid  = element_blank(),
    panel.background = element_blank()) +
  labs(title='', col='Status')
```

Graph key:

 - L: left-handedness, the exposure
 - A: ability, unobserved
 - W: winning a UFC match, the outcome
 - Q: qualification for UFC; this is required for winning a UFC match

Many readers will notice that by considering only fighters who have qualified for UFC, Q, we are here [conditioning on a collider](https://pmc.ncbi.nlm.nih.gov/articles/PMC2846442/).

Although the solution set (and the paper) treat the abilities of each fighter as purely unobserved, we are actually given some limited information on fighter abilities: since some of the fighters fight in multiple matches, we can include this information in the model with a Bradley-Terrey model:

## Bradley-Terry models

Wikipedia defines a [Bradley-Terry model](https://en.wikipedia.org/wiki/Bradley%E2%80%93Terry_model) as:

$$\Pr(i > j) = \frac{a_i}{a_i + a_j}$$

Where

 - $\Pr(i > j)$ is the probability of a pairwise comparison of element i vs element j is true; e.g. in an MMA match between fighter i and fighter j, fighter i wins
 - $a_i = e^{\beta_i}$, where $\beta_i$ is a real value assigned to element i; e.g. the skill of fighter i
 
When transformed to the log-odds scale we get:

$$\operatorname{logit}(\Pr(i > j)) = \beta_i - \beta_j$$

It seems that this class of models describes the UFClefties question.  Andy Milne describes how to parameterize this class of models for `brms` in this [stan forum post](https://discourse.mc-stan.org/t/bradley-terry-model-in-brms-or-rstanarm-lme4-notation/28932/3); essentially we need a $\beta$ variable for each fighter with a value of 1 if the fighter won the match, a value of-1 if the fighter lost, and a value of 0 otherwise.  Adding handedness gives the following model

$$\operatorname{logit}(\Pr(i > j)) = L + \beta_i - \beta_j$$

Where $L$ denotes the 'lefty advantage' and encodes the [diasteriomeric](https://en.wikipedia.org/wiki/Diastereomer) relation of the match; $L$ takes the following values:

 - 0 when a righty fights a righty or a lefty fights a lefty
 - 1 when a lefty wins over a righty
 - -1 when a righty wins over a lefty
 
Are we actually going to get much info on fighter ability from going about it this way?  Let's poke a bit at the data to find out.

## Exploratory

Let's graph the fights in the dataset with each fighter as a vertex, and each fight as an edge:

```{r fight_graph}
#| column: screen
#| out-width: 100%
#| lightbox: true             # https://quarto.org/docs/output-formats/html-lightbox-figures.html

coords <- read.csv("coords.csv") |>
# coords <- read.csv("posts/UFC/coords.csv") |>
  select(-X) |>
  as.matrix()

# graph
g <- UFClefties |>
  mutate(
    winner = if_else(fighter1.win == 1, fighter1, fighter2),
    loser = if_else(fighter1.win == 0, fighter1, fighter2)) |>
  select(winner, loser) |>
  graph_from_data_frame(
    directed = TRUE,
    vertices = fighters)

# plot
g |> plot(
  vertex.size = 5,
  vertex.label.cex = 0.2,
  edge.arrow.size = 0.2,
  layout = coords)

#cleanup
rm(coords)
```

That might be a bit difficult to see, depending on your screen width; here's a tabulation of cluster size (the number of fighters connected to one another by any number of fights):

```{r connected_table}
g.components <- components(g, mode = "weak")
fighters$connected <- g.components$membership

(fighters <- fighters |>
  group_by(connected) |>
  mutate(connected_N = n()) |>
  ungroup()) |>
  group_by(connected_N) |>
  summarize(fighters = n()) |>
  ungroup() |>
  select(connected_N, fighters) |>
  mutate(cumulative = cumsum(fighters))
```

A good chunk of the fighters are unfortunately connected to only a few, and about a fifth are in clusters of size 2; it does look like there's some good info in the graph structure though.  Let's graph the largest connected group:

```{r 92_graph}
#| column: screen
#| out-width: 100%
#| lightbox: true             # https://quarto.org/docs/output-formats/html-lightbox-figures.html

g92 <- UFClefties |>
  mutate(
    winner = if_else(fighter1.win == 1, fighter1, fighter2),
    loser = if_else(fighter1.win == 0, fighter1, fighter2)) |>
  select(winner, loser) |>
  graph_from_data_frame(
    directed = TRUE,
    vertices = fighters) |>
  delete_vertices(fighters[fighters$connected_N != 92,]$fighter)
g92 |> plot(
  vertex.size = 5,
  vertex.label.cex = 0.2,
  edge.arrow.size = 0.2,
  # margin = -0.25,
  # asp = 0.5,
  # canvas.width = 1000,
  # canvas.height = 500,
  layout = layout_with_kk(
    g92, 
    dim = 2,
    maxiter = 500*vcount(g)))
```

There's plenty more exploration to do, and this is left as an exercise for the reader.

## The Model

### Dataset

The dataset includes 244 fights, and a Bradley-Terry model requires a variable for each fight; so here's the code I used to MUNGE the data:

```{r data}
#| code-fold: show

# MUNGE the data:
d <- UFClefties |>
  mutate(
    Y = 1L,                             # response var; brms likes it this way
    # diasteriomeric description of the fight; lefty win/lose/0:
    L = as.integer((fighter1.lefty - fighter2.lefty) * (2*fighter1.win - 1)),
    fight = row_number(),               # helps pivot_wider keep track of things
    winner = if_else(fighter1.win ==1, fighter1, fighter2),
    loser = if_else(fighter1.win ==0, fighter1, fighter2)) |>
  select(Y, L, fight, winner, loser)    # gets rid of unnecessary vars

# Prepare the Bradley-Terry vars (MUNGE again):
d <- rbind(
  d |>
    mutate(fighter = winner,
           outcome = 1L) |>
    select(Y, L, fight, fighter, outcome),
  d |>
    mutate(fighter = loser,
           outcome = -1L) |>
    select(Y, L, fight, fighter, outcome)) |> 
  pivot_wider(
    id_cols = c(Y, L, fight),
    names_from = fighter,
    names_prefix = "f",
    names_sort = TRUE,
    values_from = outcome,
    values_fill = 0L) |>
  select(-fight)
```

Here's a peek of a small chunk of the fully-MUNGE-ed dataset:

```{r head_data}
#| code-fold: show
head(d[56:65,1:10], n = 10)
```

### brms Models

To start off with, here's the model from the paper and the solution set:

```{r model}
#| code-fold: show
#| output: false

model <- brm(
  formula = bf(Y ~ 0 + L, family = bernoulli),
  prior = prior(normal(0, 1), class = b),
  data = d, 
  chains = chains,
  iter = iter,
  warmup = warmup,
  seed = (seed <- as.integer(Sys.time())))
```

The seed used is `r seed`.

To specify the Bradley-Terrey model you might worry that a large formula would be needed; fortunately brms follows the standard [Wilkinson formula](https://academic.oup.com/jrsssc/article/22/3/392/6882728?login=false) syntax (e.g. that of the R stats package, nlme, the Bayesian python package bambi, etc), so instead of writing out all of the variables in the formula we can specify the formula with `bf(Y ~ 0 + ., family = bernoulli)`, where `.` denotes the sum of all variables (other than the response variable, `Y`); here `0` tells to brms not to automatically add an intercept to the model.

```{r BT_model}
#| code-fold: show
#| output: false

BT_model <- brm(
  formula = bf(Y ~ 0 + ., family = bernoulli),
  prior = prior(normal(0, 1), class = b),
  data = d, 
  chains = chains,
  iter = iter,
  warmup = warmup,
  seed = (seed <- as.integer(Sys.time())))
```

The seed used for this one is `r seed`.  For fun, let's re-run the model using only the largest cluster:

```{r BT92_model}
#| code-fold: show
#| output: false

d.92 <- d[UFClefties$fighter1 %in% (as.integer(fighters$connected_N == 92) * 1:244),]

BT92_model <- brm(
  formula = bf(Y ~ 0 + ., family = bernoulli),
  prior = prior(normal(0, 1), class = b),
  data = d.92, 
  chains = chains,
  iter = iter,
  warmup = warmup,
  seed = (seed <- as.integer(Sys.time())))
```

The seed used is `r seed`.

Trace and trank plots look fine for these models; they will not be reproduced here.

## Results

### Posterior probabilities

Here's a tabulation of the the posterior probabilities with 89% HDI:

```{r HDPI_table}
#| warning: false

(draws <- (draws.wide <- cbind(
  # Prior = rnorm(n = chains*(iter-warmup), mean = 0L, sd = 1L),
  model |> 
    as_draws_df() |>
    select(b_L) |>
    rename(Base_Model = b_L),
  BT_model |> 
    as_draws_df() |>
    select(b_L) |>
    rename(BT_Model = b_L),
  BT92_model |> 
    as_draws_df() |>
    select(b_L) |>
    rename(BT92_Model = b_L))) |>
  pivot_longer(cols = everything()) |>  # pivots all columns
  rename(Model = name)) |>
  group_by(Model) |>
  mean_hdi(value, .width = .89)
```

And the posterior predictive plot for the lefty advantage:

```{r posterior_pred_plot}
#| warning: false

draws |>
  ggplot(aes(x = value, fill = Model)) +
  stat_function(
    fun = dnorm, 
    geom = "area",
    color = "black", 
    fill = rgb(0.9, 0.5, 0, 0.1),
    alpha = 0.3) +
  geom_density(
    alpha = 0.5,
    from = -2.1,
    to = 2.1,
    n = 2048) +
  xlab("Lefty advantage") +
  ylab("Probability Density") +
  theme(                                            # add x & y lines
    panel.grid  = element_blank(),
    panel.background = element_blank(),             # remove gray background
    axis.line.x = element_line(colour = 'black', linewidth = 0.5, linetype='solid'),
    axis.line.y = element_line(colour = 'black', linewidth = 0.5, linetype='solid'),
    legend.position = "none") +
  scale_x_continuous(limits = c(-1.5, 1.5), expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, 1.7), expand = c(0, 0))
```

Plot key:

 - Yellow: normal prior
 - Red: base model
 - Green: full Bradley-Terrey model (all data points)
 - Blue: sub-Bradley-Terrey model (including only the 92 inter-connected fighters)
 
Interestingly, looking at only the 92 inter-connected fighters deviates furthest from the assumptions baked into the prior.

## Conclusion

So in this case adding information on fighter ability didn't affect the posterior distribution much, and probably didn't contain much information given the small size of the clusters, and low average number of fights for each fighter, but it was an interesting exercise.   Some parting thoughts:

 - The exercise illustrated the limited [expressive power](https://en.wikipedia.org/wiki/Expressive_power_(computer_science)) of brms relative to Stan and the Rethinking package; models will be difficult to express in brms unless they can be coerced into a design matrix, as [Wilkinson notation](https://academic.oup.com/jrsssc/article/22/3/392/6882728?login=false) fundamentally is about converting a formula to a design matrix.
 - Perhaps the "footedness" of the fighters might provide more info?  This would complicate the diasteriomeric relations in each match, as each would then be a relation between four chiral centers, but perhaps the strikes from the lower body are more important than those from the upper body.

## Version History

Updated 2025-02-14:

 - inline code
 - cleanup dag
 - code folding
 - add [lightbox](https://quarto.org/docs/output-formats/html-lightbox-figures.html) to fight graphs

Updated 2025-02-10:

 - improved graphs
 - added the 92-fighter Bradley-Terrey model
 - added license info
 - increased sample draws
 - expanded conclusions
